---
title: "BBS_modelling"
output: html_document
date: "2024-06-06"
---

#Bring in the hydro indices
```{r}
lake_indices <- read.csv("D://izzyj/rq1_analysis/lake_indices.csv")
basin_indices <- read.csv("D://izzyj/rq1_analysis/basin_indices.csv")

write_rds(lake_indices, "D://izzyj/rq1_analysis/lake_indices.rds")
write_rds(basin_indices, "D://izzyj/rq1_analysis/basin_indices.rds")
```

##Get covariate data in order
Set strata names.
```{r}
lakes_basins <- read.csv("D:/izzyj/rq1_analysis/lakes_basins.csv") #key matching lakes to basins

lakes_basins <- lakes_basins %>%
  select(lake_name, basin_name)

lake_indices <- lake_indices %>%
  left_join(lakes_basins, by = c("Name" = "lake_name"))


# Change basin_name to strata_name
lake_indices <- lake_indices %>%
  rename(strata_name = basin_name)
```

Get strata names and strata numbers
```{r}
basins_df <- as.data.frame(basins)

# Extract unique basin names and their strata numbers from the basins shapefile
basin_strata <- basins_df %>%
  select(strata_name = strata_name, strata) # Adjust column names as necessary

# Join the strata numbers to the lake_indices_joined dataframe
lake_indices <- lake_indices %>%
  left_join(basin_strata, by = "strata_name")

lake_indices
```

```{r}
basin_indices <- basin_indices %>%
  rename(strata_name = Name)

# Join the strata numbers to the basin_indices dataframe
basin_indices <- basin_indices %>%
  left_join(basin_strata, by = "strata_name")

basin_indices
```
Put all colnames into lower case
```{r}
colnames(lake_indices) <- tolower(colnames(lake_indices))
colnames(basin_indices) <- tolower(colnames(basin_indices))
```

Set types
```{r}
lake_indices$name <- as.character(lake_indices$name)
lake_indices$class <- as.factor(lake_indices$class)
lake_indices$indice <- as.factor(lake_indices$indice)
lake_indices$value <- as.numeric(lake_indices$value)

basin_indices$class <- as.factor(basin_indices$class)
basin_indices$indice <- as.factor(basin_indices$indice)
basin_indices$value <- as.numeric(basin_indices$value)
```




Split them out into separate covariate dataframes, index by index and class by class. 
```{r}
head(lake_indices)
unique(lake_indices$indice)
```

Create a list of dataframes for each class and Index.
```{r}
# Check the structure of lake_indices to ensure the columns 'class' and 'Indice' are present
str(lake_indices)

# Create a list of data frames for each unique combination of 'class' and 'Indice'
lake_indices_list <- lake_indices %>%
  group_by(class, indice) %>%
  group_split()

# Name each dataframe in the list for easy reference
names(lake_indices_list) <- lake_indices %>%
  group_by(class, indice) %>%
  group_keys() %>%
  mutate(name = paste(class, indice, sep = "_")) %>%
  pull(name)

# Check the names of the data frames created
names(lake_indices_list)

```
```{r}
# Check the structure of lake_indices to ensure the columns 'class' and 'Indice' are present
str(basin_indices)

# Create a list of data frames for each unique combination of 'class' and 'Indice'
basin_indices_list <- basin_indices %>%
  group_by(class, indice) %>%
  group_split()

# Name each dataframe in the list for easy reference
names(basin_indices_list) <- basin_indices %>%
  group_by(class, indice) %>%
  group_keys() %>%
  mutate(name = paste(class, indice, sep = "_")) %>%
  pull(name)

# Check the names of the data frames created
names(basin_indices_list)

```
Make them wide for the covariate matrix (strata name, year 1, year 2, year 3...)

```{r}
get_suffix <- function(name) {
  if (grepl("_mean$", name) || grepl("_median$", name)) {
    return("_mean")
  } else if (grepl("_max$", name) || grepl("_min$", name)) {
    return("_max_min")
  } else if (grepl("_range$", name)) {
    return("_range")
  } else {
    stop("Unknown suffix in dataframe name")
  }
}

handle_duplicates <- function(df, suffix) {
  if (suffix == "_mean") {
    df %>%
      group_by(strata_name, year) %>%
      summarise(value = mean(value, na.rm = TRUE), .groups = 'drop')
  } else if (suffix == "_max_min") {
    df %>%
      group_by(strata_name, year) %>%
      summarise(value = sum(value, na.rm = TRUE), .groups = 'drop')
  } else if (suffix == "_range") {
    df %>%
      group_by(strata_name, year) %>%
      summarise(value = max(value, na.rm = TRUE), .groups = 'drop')
  } else {
    stop("Unknown suffix")
  }
}

reshape_covariate <- function(df, suffix) {
  df %>%
    select(strata_name, year, value) %>% # Keep only necessary columns
    handle_duplicates(., suffix) %>% # Handle duplicates based on suffix
    pivot_wider(names_from = year, values_from = value) %>% # Convert to wide format
    arrange(strata_name) # Arrange by strata_name
}

# Assuming your list of data frames is named lake_indices_list
lake_indices_list_wide <- mapply(function(df, name) {
  suffix <- get_suffix(name)
  reshape_covariate(df, suffix)
}, lake_indices_list, names(lake_indices_list), SIMPLIFY = FALSE)

```

```{r}
lake_indices_list_wide[1]
```
Great, now let's just pull out the priority covariates. These are the ones we are going to model first and foremost. Then we will do lagged. Then, if we still don't fully see relationships, progress to more metrics and water classes. 
```{r}
priority_names <- c("part_cons_ANN_mean", "part_cons_ANN_range", "part_cons_SPR_mean", "part_cons_SPR_range",
                    "part_cons_FALL_mean", "part_cons_FALL_range", "perm_cons_ANN_mean", "perm_cons_ANN_range",
                    "perm_cons_SPR_mean", "perm_cons_SPR_range", "perm_cons_FALL_mean", "perm_cons_FALL_range",
                    "total_cons_ANN_mean", "total_cons_ANN_range")

priority_dataframes <- lake_indices_list_wide[names(lake_indices_list_wide) %in% priority_names]


```


##Step 1: get the model file

###REALLY ONLY NEED THIS FOR FIRST TIME OR FOR MODEL EDITS

Copy it from the package into my directory bbsbayes2_models
```{r}
#copy_model_file(model = "first_diff", model_variant = "hier", dir = "D://izzyj/rq1_analysis/bbsbayes2_models")
```
Rename this file. THIS WILL OVERWRITE EDITS MADE SO ONLY DO THIS IF YOU ARE OKAY TO START OVER.
```{r}
#file.rename("D://izzyj/rq1_analysis/bbsbayes2_models/first_diff_hier_bbs_CV_COPY.stan", "D://izzyj/rq1_analysis/bbsbayes2_models/first_diff_dswe_covariate.stan")
```
Pull it into R to edit and add chunks directly when it opens. See [https://bbsbayes.github.io/bbsBayes2/articles/advanced.html?q=covariates#example---modifying-a-model-to-include-a-covariate] 

```{r}
# Open the renamed Stan model file for editing
file.edit("D://izzyj/rq1_analysis/bbsbayes2_models/first_diff_dswe_covariate.stan")
```
Once you've made edits, save it.


###Fitting model
```{r}
species <- "Mallard"
s <- stratify(by = "basins", species = species, strata_custom = basins, return_omitted = TRUE)
strata_map <- basins

model = "first_diff"

model_variant <- "hier"

p <- prepare_data(s,
                  min_year = 2003,
                  max_year = 2019,
                  min_n_routes = 1,
                  min_max_route_years = 1,
                  min_mean_route_years = 1,
                  quiet = FALSE) 

#ps <- prepare_spatial(p,
#                      strata_map = strata_map)

# string identifying the customized model file
 cov_mod <- "D:/izzyj/rq1_analysis/bbsbayes2_models/first_diff_dswe_covariate.stan"
 
 cov_mod

 # preparing the model
pm_cov <- prepare_model(p,
                        model = model,
                        model_variant = model_variant,
                        model_file = "D:/izzyj/rq1_analysis/bbsbayes2_models/first_diff_dswe_covariate.stan")

pm_cov
```


Read in the covariate data
```{r}
#use if working directly with df list 
#cov_all <- priority_dataframes[[1]]

#otherwise, use this for sample data
cov_all <- read_rds("D://izzyj/rq1_analysis/sample_data.rds") #change to your filepath

strata_incl <- p$meta_strata # data frame with strata included in the data
years_incl <- min(p$raw_data$year) : max(p$raw_data$year) # vector of the years available
```

Check out what's going on under the hood
```{r}
strata_incl
years_incl
colnames(cov_all)
```


```{r}
library(dplyr)

cov_incl <-  strata_incl %>%
  inner_join(.,cov_all,
             by = "strata_name") %>% # dropping covariates outside of the included strata
  select(matches(as.character(years_incl)),
         strata) %>% #dropping the covariates outside of the years included
  arrange(strata) %>% # sorting to match strata ordering
  select(-strata) %>% # dropping non-covariate column
  as.matrix() # formating as numerical matrix

## adding the covariate matrix to the model_data object in the
## object created by prepare_model()
pm_cov$model_data[["cov"]] <- cov_incl
pm_cov$model_data


```

```{r}

# fitting the model using run_model()
fit_cov <- run_model(pm_cov,
                 refresh = 200,
                 iter_warmup = 2000, #increasing the warmup to support more complex model
                 iter_sampling = 4000, # increasing the sampling
                 thin = 2, # small thinning to reduce output file size
                 max_treedepth = 14, # also increased to support more complex model
                 adapt_delta = 0.85,
                 output_dir = "D:/izzyj/rq1_analysis/bbsbayes2_results", #saving model in a particular directory
                 output_basename = "covariate", #saving model output with a relevant name
                 overwrite = TRUE) 

pm_cov
pm_cov?run_model

# posterior summary and convergence of the covariate effect
cov_effect <- get_summary(fit_cov, variable = "beta_cov")
```